Stage-Splitting a Modal Language

### 2014-09-29 paper/example.tex
a lot of work on section 3

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in Figure~2a.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recursing on the sublist containing the desired element.

@@

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

>>>>

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in Figure~2a.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

@@

This paper defines how to statically perform splitting transformations, like the one described above, on programs written in a simple staged language we call \lang.  \lang\ unambiguously specifies how its terms should be split
% and features a type system that ensures valid splittings exist.
% only minimal refactoring should be required to write terms within this language,
%and from there, the type system should be enough to prove that a valid splitting exists.
by identifying each with one of two {\em stages}, namely \bbone\ or \bbtwo.
Intuitively, the stage of a \lang\ term expresses \emph{when} to evaluate it---all stage-\bbone\
subexpressions are evaluated before stage-\bbtwo\ ones.  
After splitting, those parts of the \lang\ term that were identified with stage \bbone\ 
will end up in the stage \bbone\ function ({\tt qSelect1}), 
and those parts identified with stage \bbtwo\ will end up in the stage \bbtwo\ function ({\tt qSelect2}).
Typing rules for \lang\ ensure a valid splitting exists.  That is, 
within well-typed terms, information can flow from stage \bbone\ to stage \bbtwo\ portions,
but never from \bbtwo\ to \bbone.

### 2014-09-29 paper/intro.tex
a lot of work on section 3

It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).  This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).  Examples include,
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
\end{itemize}

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

>>>>

It is a common observation that some computations can be performed efficiently by using {\em staging}, wherein the execution of a program is split over two or more discrete points in time ({\em i.e.} stages).  This most often arises when we want to evaluate a multivariate function several times, with some inputs dynamic (varying across invocations) and some static (constant across invocations).  Examples include:
\begin{itemize}
\item Computing $b^e$, where $b$ varies and $e$ is constant.
\item Finding the $i$th largest element in a list, where the list is constant and $i$ varies.
\item Computing the shading of a surface in a 3D rendering, where some contributors to the final pixel color can be approximated as constant over  pixel blocks.  
\end{itemize}

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way once inputs are available.  This has previously been explored as an optimization technique for languages with fewer features (notably, no functions) or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruf\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static program transformation and is not dependent on the input data.

### 2014-09-29 paper/semantics.tex
a lot of work on section 3

As stated in \cref{sec:staging}, we use \lang,
a typed two-stage lambda calculus featuring products, sums, and isorecursive types, as language to define and analyze the splitting transform.
The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of \cite{davies96}, 
restricted to two stages and extended with general sums and recursion\footnote{This restriction to two stages is made primarily for 
simplicity of presentation.  All of these techniques could feasibly be extended to more
complicated stage systems.}.  The following subsections provide a brief overview of the type system for \lang, and highlight important differences in its semantics from this prior work.

@@

In \lang, the $\fut$ constructor encapsulates a stage \bbtwo\ type at stage~\bbone.  
But because of the two-stage assumption, $\fut$ has no meaning when used at stage~\bbtwo.
This restriction is managed by the type validity judgment, whose rules are given in \cref{fig:validTypes}. 
In short, $\Delta \vdash A \istypewor$ means that type $A$ is valid at world $w$ under the assumptions in $\Delta$.
This context is only augmented by recursive types.

The typing judgment, like type validity, is annotated with a stage.
Thus we can read $\typeswor x A$ as saying that term $x$ has type $A$ at world $w$ under the context $\Gamma$.
The typing rules are given in \cref{fig:statics}.
The staging features ($\next$, $\prev$, and $\pause$) behave as described in the previous section,
and all other features behavior in the standard way and are essentially agnostic to stage. 
Variables in the context are annotated with the stage at which they were introduced 
and can only be used at the same stage.
As one would expect, the typing judgment only produces well-formed types 
(by induction on the derivation of the typing judgment).

@@

A key attribute of the quickselect example is that stage-\bbone\ and stage-\bbtwo\ expressions are nested. 
Ordinary term evaluation eliminates outermost redexes first, 
however in the case that stage \bbone\ expressions are contained inside stage~\bbtwo\ ones 
(such as the recurive call to {\tt fexp} above), 
this strategy conflicts with the precept of staged execution: 
that all stage-\bbone\ code be evaluated before the evaluation of stage-\bbtwo\ code.

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in a
stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
monostaged language called \langTwo. Then, we perform the remainder of the evaluation with 
$\tworedsym$, an ordinary dynamic semantics for \langTwo\ (the rules for this
judgment are not shown, but they are standard).

\subsubsection{Non-Duplicating First-Stage Evaluation}

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the following example:
\begin{lstlisting}
#2 (next {fib 20}, 2+3)
\end{lstlisting}
This is a stage-\bbone\ expression of type $\rmint$; the pair inside it is a
stage-\bbone\ expression of type $(\fut\rmint)\times\rmint$. 
In this example, ${\tt fib} : \rmint \to \rmint$ is a stage-\bbtwo\ reference to the Fibonacci function.
A conventional call-by-value evaluation strategy demands that we evaluate both components of
the pair to values before we project from the pair. The problem is that
while \verb|next {fib 20}| is not a value (in the sense that additional
stage-\bbtwo\ computation steps are necessary to produce the result \verb|6765|), 
evaluating the contents of \verb|next| cannot occur as part of stage-\bbone\ evaluation.
Intuitively, the solution is to designate \verb|next {fib 20}| as a value \emph{in
stage \bbone}, even though it requires additional evaluation in stage \bbtwo.
Therefore, we evaluate the pair to
\begin{lstlisting}
#2 (next {fib 20}, 5)
\end{lstlisting}
then obtain \verb|5| as the result of projection.

Now consider a more complex example where 
stage-\bbone\ evaluation must substitute such a incompletely-evaluated
expression. The following stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {fib 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, stage-\bbone\ evaluation will not fully reduce this term, because the answer
depends on the value of \verb|fib 20|, which is not reduced to \verb|3| until
stage \bbtwo.

If we simply treat \verb|(next {fib 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the result of stage \bbone\ computation is
\begin{lstlisting} 
prev{
  next{ 
    prev{next {fib 20}} * prev{next {fib 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s eliminate the $\next$s to yield the final residual:
\begin{lstlisting} 
(fib 20) * (fib 20) * 7
\end{lstlisting}
Note how stage-\bbtwo\ evaluation of this expression will compute \verb|fib 20| twice.  
To avoid duplicating computations, we take a different approach.  Instead of
duplicating the contents of the $\next$ expression in the tuple, we bind the contained stage-\bbtwo\ expression to
a variable (here, $\mathtt{y}$) and duplicate only a reference to that variable.
This produces:
\begin{lstlisting} 
let y = fib 20 in y * y * 7
\end{lstlisting}

Achieving this behavior mechanically requires us to resolve a contradiction:
we must substitute for \texttt{x} in stage \bbone, but we cannot evaluate inside the $\next$ block within the tuple. 
Our solution is to replace the contents of the $\next$ with a new stage-\bbtwo\ variable and create an explicit substitution (shown with a $\mapsto$) binding that stage-\bbtwo\ variable to the $\next$'s old contents.  
This substitution then floats up to the top of the containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
As a convention, we render the new variable with a %stylish and fashionable
hat.  We're now free to perform the stage-\bbone~substitution for {\tt x} without duplicating stage-\bbtwo\ work.
\begin{lstlisting} 
prev {
[yhat|->fib 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the outermost $\next$, we must first partially evaluate within its body by finding all of the contained stage-\bbone~terms and reducing them. 
As a rule, these will reduce to $\next$ expressions, which the $\prev$ eliminates, leaving the variable in place:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->fib 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = fib 20 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

Thus we have evaluated all of the stage \bbone\ expressions of this program without duplicating the contents of $\next$ blocks.

@@

\input {figures/dynamics}

The goal of the dynamics is to perform the above style of first-stage evaluation automatically.
It comprises judgments: $\redonesym$ (evaluation),
$\redtwosym$ (speculation), and $\reifysym$ (reification of substitutions into let bindings).

@@

\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment is a mapping
from fresh stage \bbtwo\ variables (which may appear inside $\next$ blocks in
$v$) to residuals---in our example above, we represented this environment with
explicit substitutions. For all of the normal features of \lang\ 
(\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and
effect on (partial) values as does standard evaluation, and the final future environment is
gotten by merging the future environments of the subterms.

When the \bbone\ judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it
searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  This search process, called
\emph{speculation}, is implemented by the $\redtwosym$ judgment, which takes
a stage \bbtwo\ term to a residual.  Once the contents of the $\next$ block are
speculated into a residual ($q$), the output of $\redonesym$ is a fresh variable wrapped in a
$\next$ block ($\next~\hat y$), along with a future environment which maps that
variable to the residual ($\hat y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does
nothing but recursively speculate into every subexpression.  Once speculation
finds a $\prev$ block, it resumes stage \bbone\ evaluation of the contents, which
produces a future context and (by canonical forms) a $\next$-wrapped variable
($\next\{\hat y\}$).  The context is then reified (using the $\reifysym$
judgment) into a series of let bindings with $\mathtt{\hat y}$, stripped of
its $\next$, at the bottom.

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage-\bbone\ code. 
Thus the evaluation of that stage one code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."

@@

Returning to our {\tt fexp} example, we can speculate on the stage \bbtwo\ term
$\verb|fn b => prev{fexp(next{b},10)}|$
to get the residual
\begin{lstlisting} 
fn b : int =>
  let x0 =
    (let x1 = 
      let x2 = 1 * b
      in x2 * x2
    in x1 * x1) * b
  in x0 * x0
\end{lstlisting}

Note how the recursion causes duplicate code in the residual.

@@

The dynamics described in the previous section provide a formal description of partial evaluation.
Recall that in partial evaluation systems, we start with a multivariate function $f$ for which some input is labeled {\em static} and 
some labeled {\em dynamic}.  Once the static input is provided, partial evaluation of $f$ produces
a residual that depends on only the dynamic input.  Equationally, a partial evaluator is any $p$ such that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to it's mathematical interpretation (a la \cite{jones96}).
Here, $x$ is the static input, $y$ is the dynamic input, and $f_x$ is the residual, also called ``f specialized to x."

@@

By identifying static with stage \bbone\ and dynamic with stage \bbtwo, 
our dynamics can serve as a partial evaluator.   
Specifically, we encode $f$ as a \lang\ expression with a function type of the form $A\to\fut(B\to C)$
\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
the following higher-order function which makes the adjustment:
\begin{lstlisting} 
let adjust (f : $int * int -> $int) =
  fn (p : int) => 
    next{
      fn (b : int) => 
        prev{f (next {b}, p)}
    }
\end{lstlisting}}.
%
Here $A$ represents the static input, $B$ represents the dynamic input, and $C$ represents the output.

@@

That this sequence of evaluations is in fact staged follows from our
characterizations of partial values, residuals, and values, that $\redonesym$
outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.

>>>>

\lang\ is a typed two-stage lambda calculus with products, sums, and isorecursive types.
Its statics are adapted from the linear temporal logic given by \cite{davies96},
restricted to two stages and extended with general sums and recursion.%
\footnote{This restriction to two stages is made primarily for simplicity of
presentation. All of these techniques could feasibly be extended to more
complicated stage systems.}
%The grammar (\cref{fig:grammar}) and type system of \lang\ are an adaptation of those of 
%, as language to define and analyze the splitting transform.

%The following subsections provide a brief overview of the type system for \lang,
%and highlight important differences in its semantics from this prior work.

@@

In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
stage~\bbone\ term. To ensure that the staging features are only applied to
valid terms, we index our judgments not only by a type, but also a stage $w$.

The type validity judgment $\Delta \vdash A \istypewor$, defined in
\cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
$\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
$\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
typing judgment only produces valid types.
%(by induction on the derivation of the typing judgment).
% under the assumptions in $\Delta$. This context is only augmented by recursive types.

Except for the staging features $\next$, $\prev$, and $\pause$, the language
features are essentially agnostic to stage. Variables in the context are
annotated with the stage at which they were introduced and can only be used at
the same stage.

Note that there is no way to eliminate a term of type $\fut A$ into a stage
\bbone\ term---the only elimination form is $\prev$, which results in a stage
\bbtwo\ term. This is why well-typed terms cannot have any information flow
from stage \bbtwo\ to stage \bbone, and what makes splitting possible.

@@

In the quickselect example, stage~\bbone\ terms are contained within
stage~\bbtwo\ terms---for example, in the recursive call to {\tt qsStaged}. In
this case, if we use an ordinary evaluation strategy reducing outermost redexes
first, we would no longer have staged evaluation, as stage~\bbtwo\ code would
be evaluated before stage~\bbone\ code.

Thus, our dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
a stage \bbtwo\ term with no stage~\bbone\ content; we say this is a term in a
monostaged language called \langTwo. Then, at stage~\bbtwo, we complete
evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
rules for this judgment are not shown, but they are standard.)

\subsubsection{Non-Duplicating Stage~\bbone\ Evaluation}

To gain intuition about the challenges of implementing this staged dynamic
semantics, consider the program:
\begin{lstlisting}
#2 (next {f 20}, 2+3)
\end{lstlisting}
This is a stage~\bbone\ expression of type $\rmint$; the pair inside it is a
stage~\bbone\ expression of type $(\fut\rmint)\times\rmint$; and ${\tt f} :
\rmint \to \rmint$ is an expensive stage~\bbtwo\ function which we would like
to avoid evaluating. A conventional call-by-value semantics would fully
evaluate both components of the pair before projecting the second component.
The problem is that, while \verb|next {f 20}| is not a value (in the sense that
additional stage-\bbtwo\ computation steps are necessary to produce a numeral),
evaluating the contents of \verb|next| cannot occur as part of stage~\bbone\
evaluation.

Intuitively, the solution is to designate \verb|next {f 20}| as a value
\emph{in stage \bbone}, even though it requires additional evaluation in stage
\bbtwo. Therefore, the pair evaluates to
\begin{lstlisting}
#2 (next {f 20}, 5)
\end{lstlisting}
and the projection, in turn, immediately evaluates to \verb|5|.

Now consider a more complex example where stage~\bbone\ evaluation must
substitute such a incompletely-evaluated expression. The following
stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {f 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, this term does not fully reduce at stage~\bbone, because it depends on
the value of \verb|f 20|, which is not reduced until stage \bbtwo.

If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the stage \bbone\ result would be
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s and $\next$s cancel to yield the residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20|
twice! To avoid this duplication, instead of treating \verb|next {f 20}| as a
value and substituting it into the body of the \verb|let|, we instead bind
\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
variable.

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->f 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[yhat|->f 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the large $\next$, we must first \emph{speculate} inside of it, to
evaluate all the stage~\bbone\ terms contained within $\prev$s. These
necessarily reduce to $\next$ expressions which cancel the $\prev$s, resulting
in:
\begin{lstlisting} 
prev {
[yhat|->f 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->f 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = f 20 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

This concludes stage \bbone\ evaluation of the program---we have reduced all
\bbone\ redexes, resulting in a monostaged term in \langTwo. Stage \bbtwo\
evaluation, $\tworedsym$, then reduces this to a numeral.

@@

\input{figures/dynamics}

Our dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above, using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(speculation), and $\reifysym$ (reification of substitutions into let
bindings).

@@

%\item Values ($\valsym$s) are \langTwo\ terms which are fully evaluated; these
%are the results of a computation after both stages have been completed. In the
%example above, the term evaluates to \verb|63|.
\end{itemize}

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment implements
the explicit substitutions in our example---it maps fresh stage \bbtwo\
variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang\ (\ref{fig:diaSemanticsCore}),
$\redonesym$ is essentially standard call-by-value evaluation, and gathers the
subterms' future environments into a single one.

At $\next~e$, in contrast, $\redonesym$ must search through $e$ for any stage
\bbone\ subexpressions, and evaluate them.  (See \ref{fig:diaSemanticsNP}.)
This search process, called \emph{speculation}, is implemented by the
$\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.

For the most part (\ref{fig:diaSemanticsSpec}), speculation simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a future context $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. We then reify $\xi$ using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.

Once speculation reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a future
environment which maps that variable to the residual ($\hat y \mapsto q$).

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."

@@

@@

These dynamics allow us to define a partial evaluator.
Recall that partial evaluators operate on multivariate functions $f$, whose
inputs are each labeled either {\em static} or {\em dynamic}. Once the static
inputs are provided, partial evaluation of $f$ produces a residual depending on
only the dynamic input. Mathematically, a partial evaluator is any $p$ such
that
\[
	\forall f,x. \exists f_x. [p(f,x) = f_x \text{ and } \forall y.\llbracket f \rrbracket(x,y)=\llbracket f_x \rrbracket(y)]
\]
where $\llbracket \cdot \rrbracket$ translates the text of a function to its
mathematical interpretation (a la \cite{jones96}). Here, $x$ is the static
input, $y$ is the dynamic input, and $f_x$ is the residual, that is, \emph{$f$
specialized to $x$}.

@@

We define a partial evaluator for \lang\ by identifying static with stage
\bbone\ and dynamic with stage \bbtwo.  Specifically, we encode $f$ as a \lang\
expression with a function type of the form $A\to\fut(B\to C)$.%
%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
%the following higher-order function which makes the adjustment:
%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%  fn (p : int) => 
%    next{
%      fn (b : int) => 
%        prev{f (next {b}, p)}
%    }
%\end{lstlisting}}
%
Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.

@@

That this sequence of evaluations is in fact staged follows from our
characterizations of partial values and residuals, that $\redonesym$
outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.

For example, to specialize quickselect to a particular list, 
\begin{lstlisting}
qsStaged [5,2,7,4,1]
\end{lstlisting}
\TODO finish example
